The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-03-09 pkg/yuimadocs/inst/doc/JSS/article-new.Rnw
update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@234 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% almost as usual
\author{Alexandre Brouste\\University of Le Mans \And 
        Masaaki Fukasawa\\Osaka University\And
        Hideitsu Hino\\Waseda University\And
        Stefano M. Iacus\\University of Milan \AND 	
        Kengo Kamatani\\The University of Tokyo \And
        Yuta Koike\\The University of Tokyo\And
        Hiroki Masuda\\Kyushu University \And
        Ryosuke Nomura\\The University of Tokyo\AND
        Teppei Ogihara\\Osaka University \And
        Yasutaka Shimuzu\\Osaka University \And
        Masayuki Uchida\\Osaka University \And
        Nakahiro Yoshida\\The University of Tokyo 
        }
\title{The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations}

@@

\subsection{How to Obtain the Package}
The \pkg{yuima} package is hosted on R-Forge and the web page
of the Project is  \url{http://r-forge.r-project.org/projects/yuima}.
The  R-Forge page contains the latest development version, and stable
version of the package will also be available through CRAN.
Development versions of the package are not supposed to be stable or functional, thus
the occasional user should consider to install the stable version first.
The package can be installed from R-Forge using
\code{install.packages("yuima",repos="http://R-Forge.R-project.org")}.

@@

\subsection{Multidimensional Processes}\label{sec:multi}
Next is an example of a system of two stochastic differential equations for the couple $(X_{t,1}, X_{t,2})$ driven by three independent Brownian motions $(W_{t,1}, W_{t,2}, W_{t,3})$
$$
\begin{aligned}
\de X_{1,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
but this has to be organized into matrix form with a vector of drift expressions and a diffusion matrix
$$
\left(\begin{array}{c}\de X_{1,t} \\\de X_{2,t}\end{array}\right)=
\left(\begin{array}{c} -3 X_{1,t} \\  -X_{1,t} - 2X_{2,t}\end{array}\right)\de t +
\left(\begin{array}{ccc}1 & 0 & X_{2,t} \\X_{1,t} & 3 & 0\end{array}\right)
\left(\begin{array}{c}\de W_{t,1} \\ \de W_{t,2} \\ \de W_{t,3}\end{array}\right)
$$
For this system it now necessary to instruct \pkg{yuima} about the state variable on bot the left-hand side of the equation and the right-hand side of the equation, i.e. the \code{solve.variable}.
<<echo=TRUE, print=FALSE,results=hide>>=
sol <- c("x1","x2") # variable for numerical solution
a <- c("-3*x1","-x1-2*x2")   # drift vector 
b <- matrix(c("1","x1","0","3","x2","0"),2,3)  #  diffusion matrix
mod3 <- setModel(drift = a, diffusion = b, solve.variable = sol)
@
Looking at the structure of the \code{noise.number} slot in \code{mod3}, one can see that this is now set to 3 which is taken as the number of columns of the diffusion matrix.
Again, this model can be easily simulated
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod3)
plot(X, plot.type="single",lty=1:2)
@

@@

Assume that we want to simulate this model
$$
\begin{aligned}
\de X^1_t &= -\theta X^1_t \de t + \de W^1_t + X^2_t \de W^3_t\\
\de X^2_t &= -(X^1_t + \gamma X^2_t) \de t + X^1_t \de W^1_t + \delta \de W^2_t
\end{aligned}
$$
Now we prepare the model using the \code{setModel} constructor function
<<setModel1>>=
sol <- c("x1","x2") # variable for numerical solution
b <- c("-theta*x1","-x1-gamma*x2")   # drift vector 
s <- matrix(c("1","x1","0","delta","x2","0"),2,3)  #  diffusion matrix
mymod <- setModel(drift = b, diffusion = s, solve.variable = sol)
@
Suppose now that we want to simulate the process on a regular grid on the interval $[0,3]$ and $n=3000$ observations. We can prepare the sampling structure as follows
<<>>=
samp <- setSampling(Terminal=3, n=3000)
@
and look inside it
<<>>=
str(samp)
@
As seen from the output, the sampling structure is quite rich and we will show how to specify some of the slots later on.
We now simulate this process specifying the \code{sampling} argument to \code{simulate}
<<>>=
set.seed(123)
X2 <- simulate(mymod, sampling=samp)
@
now the sampling structure is recorded along with the data in the \code{yuima} object \code{X2}
<<>>=
str(X2@sampling)
@
\subsubsection{Subsampling}
The sampling structure can be used to operate subsampling. Next example shows how to perform Poisson random sampling, with two independent Poisson processes, one per coordinate of \code{X2}.
<<sub1>>=
newsamp <- setSampling(
random=list(rdist=c( function(x) rexp(x, rate=10), 
function(x) rexp(x, rate=20))) )
str(newsamp)
@
In the above we have specified two independent exponential  distributions to represent Poisson arrival times. Notice that the slot \code{regular} is now set to \code{FALSE}.
Now we subsample the original trajectory of \code{X2} using the \code{subsampling} function
<<sub2,fig=TRUE,width=9,height=5>>=
newdata <- subsampling(X2, sampling=newsamp)
plot(X2,plot.type="single", lty=c(1,3),ylab="X2")
points(get.zoo.data(newdata)[[1]],col="red")
points(get.zoo.data(newdata)[[2]],col="green",pch=18)
@

@@

As mentioned above, 
Monte Carlo methods require a huge number of simulations to get the desired accuracy of the
calculation of the expectation, while the asymptotic expansion of $F^{(\ve)}$ gives very fast 
and accurate approximation by analytic formulation.
The \pkg{yuima}  package provides functions to construct the functional $F^{(\ve)}$, and automatic asymptotic expansion based on the Malliavin calculus.  starting from a  \pkg{yuima} object. 
This asymptotic expansion approach to option pricing was proposed in \citep{Yoshida92b} 
and many related papers are available today.

@@

\subsection{Asymptotic Expansion for General Stochastic Processes}
Of course, \pkg{yuima} approach is more general in that the above Levy approximation only holds when the process $X_t$ is the geometric Brownian motion. We now give an example when the underlying process $X_t$ is the following CIR model
$$
\de X_t =  0.9 X_t \de t + \ep \sqrt{X_t} \de W_t, X0 = 1
$$
and we calculate the asymptotic expansion of a european call option with strike price $K=10$ for $\ep=0.4$.

>>>>

%% almost as usual
\author{Alexandre Brouste\\University of Le Mans \And 
        Masaaki Fukasawa\\Osaka University\And
        Hideitsu Hino\\Waseda University\And
        Stefano M. Iacus\\University of Milan \AND 	
        Kengo Kamatani\\University of Tokyo \And
        Yuta Koike\\University of Tokyo\And
        Hiroki Masuda\\Kyushu University \And
        Ryosuke Nomura\\University of Tokyo\AND
        Teppei Ogihara\\Osaka University \And
        Yasutaka Shimuzu\\Osaka University \And
        Masayuki Uchida\\Osaka University \And
        Nakahiro Yoshida\\University of Tokyo 
        }
\title{The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations}

@@

\subsection{How to Obtain the Package}
The \pkg{yuima} package is hosted on R-Forge and the web page
of the Project is  \url{http://r-forge.r-project.org/projects/yuima}.
The  R-Forge page contains the latest development version, and stable
version of the package will also be available through CRAN.
Development versions of the package are not supposed to be stable or functional, thus
the occasional user should consider to install the stable version first.
The package can be installed from R-Forge using
\code{install.packages("yuima",repos="http://R-Forge.R-project.org")}.
If, the R-Forge system does not provide binary builds of the \pkg{yuima} package, the
user can also try 
\code{install.packages("yuima",repos="http://R-Forge.R-project.org", type="source")}.

@@

\subsection{Multidimensional Processes}\label{sec:multi}
Next is an example of a system of two stochastic differential equations for the couple $(X_{1,t}, X_{2,t})$ driven by three independent Brownian motions $(W_{1,t}, W_{2,t}, W_{3,t})$
$$
\begin{aligned}
\de X_{1,t} &= -3 X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + 2 X_{2,t}) \de t + X_{1,t} \de W_{1,t} + 3 \de W_{2,t}
\end{aligned}
$$
but this has to be organized into matrix form with a vector of drift expressions and a diffusion matrix
$$
\left(\begin{array}{c}\de X_{1,t} \\\de X_{2,t}\end{array}\right)=
\left(\begin{array}{c} -3 X_{1,t} \\  -X_{1,t} - 2X_{2,t}\end{array}\right)\de t +
\left(\begin{array}{ccc}1 & 0 & X_{2,t} \\X_{1,t} & 3 & 0\end{array}\right)
\left(\begin{array}{c}\de W_{1,t} \\ \de W_{2,t} \\ \de W_{3,t}\end{array}\right)
$$
For this system it now necessary to instruct \pkg{yuima} about the state variable on bot the left-hand side of the equation and the right-hand side of the equation, i.e. the \code{solve.variable}.
<<echo=TRUE, print=FALSE,results=hide>>=
sol <- c("x1","x2") # variable for numerical solution
a <- c("-3*x1","-x1-2*x2")   # drift vector 
b <- matrix(c("1","x1","0","3","x2","0"),2,3)  #  diffusion matrix
mod3 <- setModel(drift = a, diffusion = b, solve.variable = sol)
@
Looking at the structure of the \code{noise.number} slot in \code{mod3}, one can see that this is now set to 3 which is taken as the number of columns of the diffusion matrix.
Again, this model can be easily simulated
<<echo=TRUE, print=FALSE,fig=TRUE,width=9,height=4,results=hide>>=
set.seed(123)
X <- simulate(mod3)
plot(X, plot.type="single",lty=1:2)
@

@@

Assume that we want to simulate this model
$$
\begin{aligned}
\de X_{1,t} &= -\theta X_{1,t} \de t + \de W_{1,t} + X_{2,t} \de W_{3,t}\\
\de X_{2,t} &= -(X_{1,t} + \gamma X_{2,t}) \de t + X_{1,t} \de W_{1,t} + \delta \de W_{2,t}
\end{aligned}
$$
Now we prepare the model using the \code{setModel} constructor function
<<setModel1>>=
sol <- c("x1","x2") # variable for numerical solution
b <- c("-theta*x1","-x1-gamma*x2")   # drift vector 
s <- matrix(c("1","x1","0","delta","x2","0"),2,3)  #  diffusion matrix
mymod <- setModel(drift = b, diffusion = s, solve.variable = sol)
@
Suppose now that we want to simulate the process on a regular grid on the interval $[0,3]$ and $n=3000$ observations. We can prepare the sampling structure as follows
<<>>=
samp <- setSampling(Terminal=3, n=3000)
@
and look inside it
<<>>=
str(samp)
@
As seen from the output, the sampling structure is quite rich and we will show how to specify some of the slots later on.
We now simulate this process specifying the \code{sampling} argument to \code{simulate}
<<>>=
set.seed(123)
X2 <- simulate(mymod, sampling=samp)
@
now the sampling structure is recorded along with the data in the \code{yuima} object \code{X2}
<<>>=
str(X2@sampling)
@
\subsubsection{Subsampling}
The sampling structure can be used to operate subsampling. Next example shows how to perform Poisson random sampling, with two independent Poisson processes, one per coordinate of \code{X2}.
<<sub1>>=
newsamp <- setSampling(
random=list(rdist=c( function(x) rexp(x, rate=10), 
function(x) rexp(x, rate=20))) )
str(newsamp)
@
In the above we have specified two independent exponential  distributions to represent Poisson arrival times. Notice that the slot \code{regular} is now set to \code{FALSE}.
Now we subsample the original trajectory of \code{X2} using the \code{subsampling} function
<<sub2,fig=TRUE,width=9,height=5>>=
newdata <- subsampling(X2, sampling=newsamp)
plot(X2,plot.type="single", lty=c(1,3),ylab="X2")
points(get.zoo.data(newdata)[[1]],col="red")
points(get.zoo.data(newdata)[[2]],col="green",pch=18)
@

@@

As mentioned above, 
Monte Carlo methods require a huge number of simulations to get the desired accuracy of the
calculation of the expectation, while the asymptotic expansion of $F^{(\ve)}$ gives very fast 
and accurate approximation by analytic formulation.
The \pkg{yuima}  package provides functions to construct the functional $F^{(\ve)}$, and automatic asymptotic expansion based on the Malliavin calculus.  starting from a  \pkg{yuima} object. 
This asymptotic expansion approach to option pricing was proposed in early 1990's 
( \citep{Yoshida92b}, \citep{Takahashi99}, \citep{Kunitomo01}), and 
many related papers are available today.

@@

\subsubsection{Asymptotic Expansion for General Stochastic Processes}
Of course, \pkg{yuima} approach is more general in that the above Levy approximation only holds when the process $X_t$ is the geometric Brownian motion. We now give an example when the underlying process $X_t$ is the following CIR model
$$
\de X_t =  0.9 X_t \de t + \ep \sqrt{X_t} \de W_t, X0 = 1
$$
and we calculate the asymptotic expansion of a european call option with strike price $K=10$ for $\ep=0.4$.
